---
title: 'Population Trends by Region'
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---
0.  The World Bank publishes data sets which track the population of countries and pre-defined regions over a period of 55 years.
    The regional data is classified into socio-economic catagories so comparing this data may tell us something about trends
    underlying those catagories.

  http://data.worldbank.org/indicator/SP.POP.TOTL
  
The World Bank defines total population as follows . . .

Total population is based on the de facto definition of population, which counts all residents regardless of legal status or       citizenship--except for refugees not permanently settled in the country of asylum, who are generally considered part of the 
population of their country of origin. The values shown are midyear estimates.

This rudimentary analysis looks at the 26 regional data sets contained in the World Bank's super-set, presents a historgram
of decreasing population counts by region and scatterplots of each region's popuation trend faceted by region.



```{r load-pkgs, eval=TRUE, echo=FALSE, message=FALSE}
library('dplyr')
library('tidyr')
library("knitr")
library("ggplot2")
```


1. Load Data Frame from file
```{r load-text-by-line, eval=TRUE}
l <- read.csv(
              "/Users/scottkarr/IS607Spring2016/project2/more/population.csv",
              sep=",",
              na.strings = "",
              blank.lines.skip = TRUE,
              #col.names = c("Quintile", "West",  "South", "Midwest","Northeast", "US Overall"),
              stringsAsFactors=FALSE
    )
df = data.frame(l)
names(df) <- gsub("X", "", names(df))
#kable(head(df), align = 'l')
```

2.  Tidy data
```{r tidy-data, eval=TRUE}
# filter columns
df <- df %>% 
  select(Country.Name, Country.Code, 5:ncol(df))
# gather morphs data from wide to long format
df_tidy <- df %>% 
  gather(Year, Population, -Country.Name, -Country.Code) %>%
  arrange(Country.Name, Year, Population)
# make year and population numerics
df_tidy$Year <- as.numeric(df_tidy$Year)
df_tidy$Population <- as.numeric(df_tidy$Population)
# remove na rows
df_tidy <- df_tidy %>% na.omit()
# present data nicely
kable(head(df_tidy), align = 'l')
```

3.  Analyze data - let's look only at regional population trends
```{r analysis-domain, eval=TRUE}
# filter only on regional population data sets
df_regions <- 
  filter(
    df_tidy %>% select(Country.Name, Country.Code) %>% distinct()
    , 
      grepl(' all', Country.Name) | 
      grepl('states', Country.Name) | 
      grepl('members', Country.Name) |     
      grepl('poor', Country.Name) |         
      grepl('conflict', Country.Name) |             
      grepl(':', Country.Name) |             
      grepl('income', Country.Name) |
      grepl('only', Country.Name) |
      grepl('small', Country.Name)    
  )
# display key codes for regions
kable(df_regions, align = 'l')

# filter the presentation data set by these "target" regions
target <- df_regions$Country.Code
df_regions <- df_tidy %>% filter (Country.Code %in% target)
```

4.  Presentation
```{r plot, eval=TRUE}
# histogram of population by regions
ggplot(df_regions) + geom_histogram(aes(x = Population))
# regional scatterplot of population by regions
ggplot(data = df_regions, aes(x = Year, y = Population)) +
  geom_point() + facet_wrap( ~ Country.Code )
```



