---
title: 'GINI Index for Regional US income'
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---
0.  The Census Bureau publishes data sets which tracks the GINI index at different levels of geographic granularity . . . such as state, metropolitan statical area and congressional district.  This exercise cleans up "untidy" contingency tables for GINI by region.  Census
data at various levels can be searched under the American Community Survey for the year being sought by under the subject of income, by
either individual or family and at a chosen geographic level  The census provides the followng tool for acquiring data sets . . .

  http://factfinder.census.gov/faces/nav/jsf/pages/guided_search.xhtml


GINI is an internationally recognized measure of income dispersion within a specified geographic area.  Income inequality has been a topical discussion in recent years and this is a key measure.


```{r load-pkgs, eval=TRUE, echo=FALSE, message=FALSE}
library('ggplot2')
library('knitr')
library('tidyr')
library('dplyr')
```

1. Load Data Frame from file
```{r load-text-by-line, eval=TRUE}
l <- read.csv(
              "/Users/scottkarr/IS607Spring2016/project2/more/GINI-2014-Region-untidy.csv",
              sep=",",
              na.strings = "",
              blank.lines.skip = TRUE,
              col.names = c("Quintile", "West",  "South", "Midwest","Northeast", "US Overall"),
              stringsAsFactors=FALSE
    )
df = data.frame(l)
```

2.  Tidy data
```{r tidy-data, eval=TRUE}
# remove extraneous rows
# derived fields can be calculated from raw data
df <- df[-c(1,7),]
# gather morphs data from wide to long format
df_tidy <- df %>% 
  gather(Region, Gini, -Quintile) %>%
  arrange(Quintile, Region, Gini)
# organize the final data sets
df_tidy <- df_tidy %>%
  select(Region, Quintile, Gini) %>%
  arrange(Region, Quintile, Gini)
# present data nicely
kable(df_tidy, align = 'l')
```


3. Analysis - Group by region and calculate statistics
```{r analyze, eval=TRUE}
df_tidy_grouped= group_by(df_tidy, Region)
df_stats <-summarise(df_tidy_grouped, mean_gini = mean(Gini), std_gini = sd(Gini))
```


4.  Presentation
```{r plot, eval=TRUE}
# histogram of population by regions
ggplot(df_tidy) + geom_histogram(aes(x = Gini))
# regional scatterplot of population by regions
ggplot(data = df_tidy, aes(x = Quintile, y = Gini)) +
  geom_point() + facet_wrap( ~ Region )
```
